An elaborate data set on human gait and the effect of mechanical perturbations

### 2014-12-05 paper.tex
Added GATK Python snippets, closes issue #24.

\usepackage{graphicx}
\usepackage{booktabs} % for table /toprule, /midrule, etc

@@

@@

\begin{figure*}
  \includegraphics{figures/frequency_analysis.pdf}
  \caption{Frequency analysis of the treadmill belt velocity input signal
                 (purple) and the recorded output velocity (blue)
                for average belt speeds of 0.8, 1.2, and 1.6 m/s, respectively}
  \label{fig:freq_analysis}
\end{figure*}

@@

We developed a command line data processing tool for common gait computations
and provide the processed data from this. The tool was developed in Pythin and centered around two main
classes, one to store a complete trial in memory and a second to compute common
gait variables of interest.

The \verb|motek.DFlowData| class collects and stores all the raw data presented
in the previous section and applies several ``cleaning'' operations to put the
data into a usable form.

\begin{itemize}
  \item Loads the meta data file into a Python dictionary.
  \item Loads the DFlow mocap module TSV file into Pandas \verb|DataFrame|.
  \item Relabels the column headers to more meaningful names if this is
    specified in the meta data.
  \item Optionally identifies the missing values in the mocap marker data and
    replaces them with \verb|numpy.nan|.
  \item Optionally interpolates the missing marker values and replaces them
    with interpolated estimates using a variety of interpolation methods.
  \item Loads the D-Flow record moudle TSV file into a Pandas \verb|DataFrame|.
  \item Extracts the events and creates a dictionary mapping the event names in
    the meta data to the events detected in the record module file.
  \item Compensates the ground reaction loads for treadmill motion.
  \item Merges the data from the mocap module and record module into one data
    frame at the maximum common constant sample rate.
  \item Extracts sections of the data bounded by the events.
  \item Saves the cleaned data to disk.
\end{itemize}

The \verb|gait.WalkingData| class is then used to compute things such as gait
landmarks (toe off and heel strike times), basic 2D inverse dynamics, and to
store the data into a Pandas \verb|Panel| with each gait cycle on the main axis
at a specified sample rate. This data is also saved to disk in HDF5 format.

Figure~\ref{fig:angle-torque-comparison} compares the mean and standard
deviation of saggital plane joint angles and torques from the perturbed gait
cycles and the unperturbed gait cycles computed from trial 20 for 120 gait
cycles. The variation in the dynamics is larger for the perturbed gait cycles.

\begin{figure*}
  \includegraphics{figures/unperturbed-perturbed-comparison.pdf}
  \caption{Mean (solid, dashed) and $2\sigma$ (shaded) ($N=120$) joint angles
    and torques from both unperturbed (purple) and perturbed (blue) gait cycles
    from trial 20.}
  \label{fig:angle-torque-comparison}
\end{figure*}

@@

\begin{figure*}
  \includegraphics{figures/unperturbed-perturbed-boxplot-comparison.pdf}
  \caption{Box plots of the average belt speed, stride frequency, and stride
    length which compare 120 gait cycles for the unperturbed (purple) and
    perturbed (blue).}
  \label{fig:gait-cycle-stats-comparison}
\end{figure*}

>>>>

% graphicx is also declared in f1000_styles
\usepackage{graphicx}

% for nice table formatting, i.e. /toprule, /midrule, etc
\usepackage{booktabs}

% to allow for \verb++ declarations in captions.
\usepackage{cprotect}

@@

% This allows Figures to be renamed as Listing, as required by F1000Research.
\usepackage{float}
\newfloat{listing}{thp}{lop}
\floatname{listing}{Listing}

@@

\begin{figure*}
  \includegraphics{figures/frequency_analysis.pdf}
  \caption{Frequency spectrum of the treadmill belt velocity input signal
    (purple) and the recorded output velocity (blue) for average belt speeds of
    0.8, 1.2, and 1.6 m/s, respectively}
  \label{fig:freq_analysis}
\end{figure*}

@@

We developed a toolkit for data processing, GaitAnalysisToolKit, for common
gait computations and provide an example processed trial to show the character
of the data. The tool was developed in Python, is dependent on the SciPy Stack,
and provides two main classes: one to do basic gait data cleaning,
\verb|DFlowData|, and a second to compute common gait variables of interest,
\verb|GaitData|.

The \verb|DFlowData| class collects and stores all the raw data presented in
the previous section and applies several ``cleaning'' operations to transform
the data into a usable form. The cleaning process follows these basic steps:

\begin{enumerate}
  \item Loads the meta data file into a Python dictionary.
  \item Loads the DFlow mocap module TSV file into Pandas \verb|DataFrame|.
  \item Relabels the column headers to more meaningful names if this is
    specified in the meta data.
  \item Optionally identifies the missing values in the mocap marker data and
    replaces them with \verb|numpy.nan|.
  \item Optionally interpolates the missing marker values and replaces them
    with interpolated estimates using a variety of interpolation methods.
  \item Loads the D-Flow record module TSV file into a Pandas \verb|DataFrame|.
  \item Extracts the events and creates a dictionary mapping the event names in
    the meta data to the events detected in the record module file.
  \item Compensates the ground reaction loads based on whether the meta data
    indicates there was treadmill motion.
  \item Merges the data from the mocap module and record module into one data
    frame at the maximum common constant sample rate.
\end{enumerate}

Once the data is cleaned there are two methods that allow you to extract the
cleaned data:

\begin{itemize}
  \item Extract sections of the data bounded by the events recorded in the
    \verb|record-XXX.txt| file.
  \item Save the cleaned data to disk.
\end{itemize}

An example of the API is given in Listing~\ref{lis:dflowdata}.

\begin{listing*}
  \begin{verbatim}
    >>> from gaitanalysis.motek import DFlowData
    >>> data = DFlowData('mocap-020.txt', 'record-020.txt', 'meta-020.yml')
    >>> mass = data.meta['subject']['mass']
    >>> data.clean()
    >>> event_df = dflow_data.extract_processed_data(event='Longitudinal Perturbation')
  \end{verbatim}
  \cprotect\caption{Python interpreter session showing how one could load a
    trial into memory, extract the subject's mass from the meta data, run the
    data cleaning process, and finally extract a Pandas \verb|DataFrame|
    containing all of the time histories for a specific event in the trial.}
  \label{lis:dflowdata}
\end{listing*}

The \verb|GaitData| class is then used to compute things such as gait landmarks
(toe off and heel strike times), basic 2D inverse dynamics, and to store the
data into a Pandas \verb|Panel| with each gait cycle on the main axis at a
specified sample rate. This data is also saved to disk in HDF5 format.

\begin{listing*}
  \begin{verbatim}
    >>> from gaitanalysis.gait import GaitData
    >>> gdata = GaitData(event_df)
    >>> gdata.inverse_dynamics_2d(left_markers, right_markers, left_loads,
    ...                           right_loads, mass, 6.0)
    >>> gdata.grf_landmarks('Right Fy', 'Left Fy', threshhold=20.0)
    >>> gdata.split_at('right')
    >>> gdata.plot_gait_cycles('Left Hip Joint Torque', mean=True)
    >>> gdata.save('gait-data.h5')
  \end{verbatim}
  \cprotect\caption{Python interpreter session showing how one could compute
    the inverse dynamics (joint angles and torques), indentify the gait
    landmarks (e.g. heelstrikes), split the data with respect to the gait
    landmarks in a Pandas \verb|Panel|, plot the mean and standard deviation of
    one time history with respect to the gait cycles, and save the data to disk.}
  \label{lis:dflowdata}
\end{listing*}

A similar workflow was used to produce Figure~\ref{fig:angle-torque-comparison}
which compares the mean and standard deviation of saggital plane joint angles
and torques from the perturbed gait cycles and the unperturbed gait cycles
computed from trial 20 for 120 gait cycles. The variation in the dynamics is
larger for the perturbed gait cycles.

\begin{figure*}
  \centering
  \includegraphics{figures/unperturbed-perturbed-comparison.pdf}
  \cprotect\caption{Mean (right: solid, left: dashed) and $3\sigma$ (shaded)
    joint angles and torques from both unperturbed (purple) and perturbed
    (blue) gait cycles from trial 20. Produced by \verb|unperturbed_perturbed_comparison.py|.}
  \label{fig:angle-torque-comparison}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics{figures/unperturbed-perturbed-boxplot-comparison.pdf}
  \cprotect\caption{Box plots of the average belt speed, stride frequency, and
    stride length which compare 120 gait cycles for the unperturbed (purple) and
    perturbed (blue). Produced by \verb|unperturbed_perturbed_comparison.py|.}
  \label{fig:gait-cycle-stats-comparison}
\end{figure*}
